---
title: "Theming and Scales"
format: html
vignette: >
  %\VignetteIndexEntry{Theming and Scales}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 3.5,
  fig.align = "center"
)
library(ggplot2)
library(mariner)
```

```{=html}
<!--
Hand-rolled tabs, because Quarto's `::: {.panel-tabset}` does not survive the
vignette build. The quarto::html vignette engine renders with `document-css:
false` and no Bootstrap, so a panel-tabset comes out as a bare class on stacked
divs with no nav, no panes and no JS -- the tabs silently become five headings
in a row. This is ~30 lines and depends on nothing.

Each panel is `::: {.mar-tab data-title="..."}` inside a `::: {.mar-tabs}`; the nav
is built from the data-title attributes at load.

The class names must NOT contain the substring "tabset". pkgdown finds tabsets
with the XPath `//div[contains(@class, 'tabset')]` -- a substring match, not a
class match -- so an earlier `.mtabset` was picked up by pkgdown's own tabset
rewriter, which replaced the first panel with a Bootstrap `tab-pane` carrying
another section's id and broke the set.
-->
<style>
.mar-tabs { margin: 1.25em 0; }
.mar-tabs-nav {
  display: flex; flex-wrap: wrap; gap: 0.15rem;
  border-bottom: 1px solid #c2c8c4; margin-bottom: 1rem;
}
.mar-tabs-nav button {
  font: inherit; background: none; border: 0; cursor: pointer;
  padding: 0.45em 0.9em; color: #6f7671;
  border-bottom: 2px solid transparent; margin-bottom: -1px;
}
.mar-tabs-nav button:hover { color: #154734; }
.mar-tabs-nav button[aria-selected="true"] {
  color: #154734; border-bottom-color: #154734; font-weight: 600;
}
.mar-tabs > .mar-tab { display: none; }
.mar-tabs > .mar-tab.is-active { display: block; }
</style>
<script>
document.addEventListener("DOMContentLoaded", function () {
  document.querySelectorAll(".mar-tabs").forEach(function (set) {
    var tabs = Array.prototype.filter.call(
      set.children, function (el) { return el.classList.contains("mar-tab"); }
    );
    if (!tabs.length) return;
    var nav = document.createElement("div");
    nav.className = "mar-tabs-nav";
    nav.setAttribute("role", "tablist");
    tabs.forEach(function (tab, i) {
      var b = document.createElement("button");
      b.type = "button";
      b.setAttribute("role", "tab");
      b.textContent = tab.getAttribute("data-title") || "Tab " + (i + 1);
      b.setAttribute("aria-selected", i === 0 ? "true" : "false");
      b.addEventListener("click", function () {
        tabs.forEach(function (t) { t.classList.remove("is-active"); });
        nav.querySelectorAll("button").forEach(function (x) {
          x.setAttribute("aria-selected", "false");
        });
        tab.classList.add("is-active");
        b.setAttribute("aria-selected", "true");
      });
      if (i === 0) tab.classList.add("is-active");
      nav.appendChild(b);
    });
    set.insertBefore(nav, tabs[0]);
  });
});
</script>
```

**mariner** ships a document theme and a matching palette for **ggplot2**.

## 1. Brand Palette and Chrome

The brand colours split in two.

* **Chrome roles**: the page background, headings, rules and link text.
* **Data series**: the colours a figure draws with. Each one clears a contrast
  threshold against the background and stays separable under colour vision
  deficiency.

`mariner_colors()` returns every colour the theme defines, by name:

```{r brand-colors}
names(mariner_colors())

mariner_colors()[c("background", "ink", "primary", "secondary", "link")]
```

### Brand Chrome Roles

```{r chrome-swatch, fig.height = 1.8, echo = FALSE}
chrome_roles  <- c("primary", "secondary", "accent-ink", "accent-rule",
                   "link", "ink", "ink-muted")
chrome_labels <- c("Primary", "Secondary", "Accent Ink", "Accent Rule",
                   "Link", "Ink", "Muted Ink")

chrome_df <- data.frame(
  Slot  = seq_along(chrome_roles),
  Role  = factor(chrome_labels, levels = chrome_labels),
  Color = unname(mariner_colors()[chrome_roles])
)

ggplot(chrome_df, aes(Slot, 1, fill = Color)) +
  geom_tile(color = mariner_colors()[["background"]],
            linewidth = 1.2, width = 0.88, height = 0.88) +
  scale_fill_identity() +
  scale_x_continuous(breaks = chrome_df$Slot, labels = chrome_df$Role,
                     position = "top", expand = c(0.02, 0.02)) +
  scale_y_continuous(breaks = NULL, expand = c(0.04, 0.04)) +
  coord_fixed() +
  labs(x = NULL, y = NULL) +
  theme_mariner() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text.y = element_blank()
  )
```

## 2. Palette Families

`mariner_pal()` covers four families:

1. **Discrete (`"discrete"`)**: categories, up to 8 classes.
2. **Sequential (`"sequential"`)**: magnitude.
3. **Ordinal (`"ordinal"`)**: ordered factors.
4. **Diverging (`"diverging"`)**: values around a neutral midpoint.

The last row below is not a family. Chart chrome is a set of *roles* from
`mariner_colors()` — the ink and rules a panel is drawn with. It is here because
a figure uses those colours without being asked.

```{r palette-swatches, fig.height = 2.6, echo = FALSE}
fams <- list(
  "Categorical (8)"  = mariner_pal(family = "discrete",   n = 8),
  "Sequential (7)"   = mariner_pal(family = "sequential", n = 7),
  "Ordinal (5)"      = mariner_pal(family = "ordinal",    n = 5),
  "Diverging (7)"    = mariner_pal(family = "diverging",  n = 7),
  "Chart chrome (5)" = unname(mariner_colors()[
    c("ink", "ink-secondary", "ink-muted", "rule-axis", "rule-grid")])
)

swatches <- do.call(rbind, lapply(names(fams), function(f) {
  data.frame(Family = f, Step = seq_along(fams[[f]]), Color = fams[[f]])
}))
swatches$Family <- factor(swatches$Family, levels = rev(names(fams)))

ggplot(swatches, aes(Step, Family, fill = Color)) +
  geom_tile(color = mariner_colors()[["background"]],
            linewidth = 1.2, height = 0.72) +
  scale_fill_identity() +
  scale_x_continuous(breaks = NULL, expand = c(0.01, 0.01)) +
  labs(x = NULL, y = NULL) +
  theme_mariner() +
  theme(
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(face = "bold")
  )
```

## 3. Scale Functions

Five scale families reach four palettes. Sequential is spent twice:
`scale_*_mariner_c()` runs it as a gradient, `scale_*_mariner_b()` cuts it into
steps. The rest map one to one — `_d()` to discrete, `_o()` to ordinal, `_div()`
to diverging. Each has a `colour` and a `fill` form, and `scale_color_*` works
everywhere `scale_colour_*` does.

One dataset throughout, fuel economy for 234 vehicle models, so a difference
between two tabs is the scale rather than a change of subject. The
[gallery PDF](../mariner-gallery.pdf) carries the same figures.

```{r scale-data}
mpg_data <- ggplot2::mpg
mpg_data$class <- factor(mpg_data$class)

# Ordinal needs an ordered variable, so displacement is cut into ranked bands.
mpg_data$size <- cut(mpg_data$displ, 3,
                     labels = c("Small", "Medium", "Large"))

class_summary <- aggregate(cbind(cty, hwy) ~ class, data = mpg_data, FUN = mean)
class_summary$delta <- class_summary$hwy - mean(mpg_data$hwy)
class_summary <- class_summary[order(class_summary$delta), ]
class_summary$class <- factor(class_summary$class, levels = class_summary$class)
```

::: {.mar-tabs}

::: {.mar-tab data-title="Discrete"}

`scale_colour_mariner_d()` — categories with no order. The observations
themselves rather than a boxplot, which would summarise them and hide how many
models sit behind each class.

```{r scale-discrete}
ggplot(mpg_data, aes(class, hwy, colour = class)) +
  geom_jitter(width = 0.22, height = 0, alpha = 0.75, size = 1.8) +
  scale_colour_mariner_d() +
  labs(x = NULL, y = "Highway mpg") +
  theme_mariner() +
  theme(legend.position = "none")
```

:::

::: {.mar-tab data-title="Continuous"}

`scale_colour_mariner_c()` — magnitude on one hue, as a smooth gradient.

```{r scale-continuous}
ggplot(mpg_data, aes(displ, hwy, colour = cty)) +
  geom_point(size = 2.2, alpha = 0.9) +
  scale_colour_mariner_c() +
  labs(x = "Engine displacement (L)", y = "Highway mpg", colour = "City mpg") +
  theme_mariner()
```

:::

::: {.mar-tab data-title="Ordinal"}

`scale_fill_mariner_o()` — a few ranked levels. The palette ramps rather than
contrasts, so the colours carry the order.

```{r scale-ordinal}
ggplot(mpg_data, aes(size, fill = size)) +
  geom_bar(width = 0.6) +
  scale_fill_mariner_o() +
  labs(x = "Engine size band", y = "Models") +
  theme_mariner() +
  theme(legend.position = "none")
```

:::

::: {.mar-tab data-title="Diverging"}

`scale_fill_mariner_div()` — a signed quantity, where the neutral step means
*no difference*. The limits are symmetric on purpose. Without that, the neutral
step lands wherever the data straddles and the scale misplaces zero.

```{r scale-diverging}
lim <- max(abs(class_summary$delta))

ggplot(class_summary, aes(class, delta, fill = delta)) +
  geom_col(width = 0.7) +
  scale_fill_mariner_div(limits = c(-lim, lim)) +
  labs(x = NULL, y = "Difference from overall mean (mpg)") +
  theme_mariner() +
  theme(legend.position = "none")
```

:::

::: {.mar-tab data-title="Binned"}

`scale_colour_mariner_b()` — the sequential palette cut into steps. Use it when
a reader has to name the band a point falls in, not just rank two points by
eye.

```{r scale-binned}
ggplot(mpg_data, aes(displ, hwy, colour = cty)) +
  geom_point(size = 2.2, alpha = 0.9) +
  scale_colour_mariner_b() +
  labs(x = "Engine displacement (L)", y = "Highway mpg", colour = "City mpg") +
  theme_mariner()
```

:::

:::

## 4. Knitr Chunk Configuration

In the setup chunk of a report, `mariner_knitr_setup()` sets the chunk options
and the ggplot2 theme:

```r
mariner_knitr_setup("pdf")
```

## 5. Document Markup and Callouts

The theme gives a report two more things. Both act on **PDF output only**: a
Lua filter in the Quarto extension carries them, so they do nothing in this HTML
vignette. The [gallery PDF](../mariner-gallery.pdf) has them typeset.

**Markup for defined terms and blocks:**

| Written | Renders as |
|---|---|
| `[formal sum]{.defn}` | a defined term — underlined italic, brand primary |
| `[basis]{.term}` | a key term — bold, brand primary |
| `[combination]{.termref}` | a reference back to one — italic, brand primary |
| `[unreduced]{.emph}` | strong emphasis — underlined bold italic |
| `[A B C]{.font-headings}` | a specimen set in the heading face |
| `[A B C]{.font-body}` | a specimen set in the body face |
| `::: {.def}` … `:::` | a definition block, indented under a rule |
| `::: {.thm}` … `:::` | a theorem block, ruled above and below |

An identifier on a `.defn` — `[support]{.defn #support}` — also plants a
cross-reference target.

Markdown `*italics*` are untouched. `.emph` is its own mark, not a restyling of
emphasis.

**Callouts.** Quarto's five — `::: {.callout-note}`, `.callout-tip`,
`.callout-warning`, `.callout-important` and `.callout-caution` — work as usual
and take the brand colours on their border and title.
