---
title: "Theming and Scales"
format: html
vignette: >
  %\VignetteIndexEntry{Theming and Scales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 3.5,
  fig.align = "center"
)
library(ggplot2)
library(mariner)
```

`mariner` provides a Baylor theme and a palette system for `ggplot2`.

## 1. Brand Palette and Chrome

Brand colours separate into chrome roles and data series palettes.

* **Chrome roles**: Page background, headings, rules, and link text.
* **Data series**: Colors selected and tested for contrast against the background and distinction under color vision deficiencies.

Inspect available brand hex codes:

```{r brand-colors}
mariner_colors()
```

### Brand Chrome Roles

```{r chrome-swatch, fig.height = 1.8}
chrome_roles  <- c("primary", "secondary", "accent-ink", "accent-rule",
                   "link", "ink", "ink-muted")
chrome_labels <- c("Primary", "Secondary", "Accent Ink", "Accent Rule",
                   "Link", "Ink", "Muted Ink")

chrome_df <- data.frame(
  Slot  = seq_along(chrome_roles),
  Role  = factor(chrome_labels, levels = chrome_labels),
  Color = unname(mariner_colors()[chrome_roles])
)

ggplot(chrome_df, aes(Slot, 1, fill = Color)) +
  geom_tile(color = mariner_colors()[["background"]],
            linewidth = 1.2, width = 0.88, height = 0.88) +
  scale_fill_identity() +
  scale_x_continuous(breaks = chrome_df$Slot, labels = chrome_df$Role,
                     position = "top", expand = c(0.02, 0.02)) +
  scale_y_continuous(breaks = NULL, expand = c(0.04, 0.04)) +
  coord_fixed() +
  labs(x = NULL, y = NULL) +
  theme_mariner() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text.y = element_blank()
  )
```

## 2. Palette Families

`mariner_pal()` generates palettes across five families:

1. **Discrete (`"discrete"`)**: Categorical data up to 8 classes.
2. **Sequential (`"sequential"`)**: Quantitative magnitude.
3. **Ordinal (`"ordinal"`)**: Ordered factors.
4. **Diverging (`"diverging"`)**: Values with a neutral midpoint.
5. **Chart chrome**: Structural grid and axis rules.

```{r palette-swatches, fig.height = 2.6}
fams <- list(
  "Categorical (8)"  = mariner_pal(family = "discrete",   n = 8),
  "Sequential (7)"   = mariner_pal(family = "sequential", n = 7),
  "Ordinal (5)"      = mariner_pal(family = "ordinal",    n = 5),
  "Diverging (7)"    = mariner_pal(family = "diverging",  n = 7),
  "Chart chrome (5)" = unname(mariner_colors()[
    c("ink", "ink-secondary", "ink-muted", "rule-axis", "rule-grid")])
)

swatches <- do.call(rbind, lapply(names(fams), function(f) {
  data.frame(Family = f, Step = seq_along(fams[[f]]), Color = fams[[f]])
}))
swatches$Family <- factor(swatches$Family, levels = rev(names(fams)))

ggplot(swatches, aes(Step, Family, fill = Color)) +
  geom_tile(color = mariner_colors()[["background"]],
            linewidth = 1.2, height = 0.72) +
  scale_fill_identity() +
  scale_x_continuous(breaks = NULL, expand = c(0.01, 0.01)) +
  labs(x = NULL, y = NULL) +
  theme_mariner() +
  theme(
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(face = "bold")
  )
```

## 3. Scale Functions

`mariner` exports scale functions for `colour` and `fill`.

### Discrete Scale: `scale_colour_mariner_d()`

```{r scale-discrete}
ggplot(mpg, aes(class, hwy, color = class)) +
  geom_jitter(width = 0.2, height = 0, size = 2) +
  scale_colour_mariner_d() +
  labs(title = "Fuel Economy by Vehicle Class",
       x = "Vehicle Class", y = "Highway MPG") +
  theme_mariner() +
  theme(legend.position = "none")
```

### Continuous Scale: `scale_colour_mariner_c()`

```{r scale-continuous}
ggplot(mpg, aes(displ, hwy, color = cty)) +
  geom_point(size = 2.5) +
  scale_colour_mariner_c() +
  labs(title = "Displacement and Highway MPG",
       x = "Engine Displacement (L)", y = "Highway MPG", color = "City MPG") +
  theme_mariner()
```

### Diverging Scale: `scale_fill_mariner_div()`

Diverging scales center on zero or a reference value with symmetric limits:

```{r scale-diverging}
class_mean <- tapply(mpg$hwy, mpg$class, mean)
class_deviation <- data.frame(
  class = names(class_mean),
  deviation = unname(class_mean) - mean(mpg$hwy)
)
class_deviation <- class_deviation[order(class_deviation$deviation), ]
class_deviation$class <- factor(class_deviation$class, levels = class_deviation$class)

div_limit <- max(abs(class_deviation$deviation)) * c(-1, 1)

ggplot(class_deviation, aes(class, deviation, fill = deviation)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_mariner_div(limits = div_limit) +
  labs(title = "Highway MPG Relative to Mean",
       x = NULL, y = "MPG Deviation from Average", fill = "Deviation") +
  theme_mariner()
```

### Ordinal Scale: `scale_fill_mariner_o()`

```{r scale-ordinal}
ggplot(diamonds, aes(x = cut, fill = cut)) +
  geom_bar(width = 0.6) +
  scale_fill_mariner_o() +
  labs(title = "Diamond Counts by Cut Quality", x = "Cut", y = "Count") +
  theme_mariner() +
  theme(legend.position = "none")
```

### Binned Scale: `scale_fill_mariner_b()`

```{r scale-binned}
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  scale_fill_mariner_b() +
  labs(title = "Old Faithful Eruption Density",
       x = "Waiting Time (min)", y = "Eruption Duration (min)", fill = "Density") +
  theme_mariner()
```

## 4. Knitr Chunk Configuration

In report `.qmd` setup chunks, `mariner_knitr_setup()` initializes options and sets the ggplot2 theme:

```r
mariner_knitr_setup("pdf")
```
